Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete

### 2019-05-23 analysis/paper/paper.Rmd
Discussion and Conclusion

Looking at Figure \@ref(fig:dimensionsboxplots), though, we can see that, through the centuries, the dimensions of the tombs varied and sometimes consistently, as with the trend toward narrower *dromoi*. The 9th-century tombs, including all but one of those seven noted above, do not particularly stand out. However, it does seem that chamber were, on average, larger in this period as well as varying more in size between tombs. Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on the construction of the larger chambers would have been limited. The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this. It appears at any rate that they were gradually declining in size, albeit with some notable outliers.

@@

ggplot()+geom_col(data=import_aorist,aes(decade,aoristic_sum),alpha=0.2)+
          geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
            theme_bw()+theme(text=element_text(family="Garamond", size=12))+
                labs(x="Date (Years BC)", y="Aoristic Sum")+
                  scale_fill_brewer(name="Origin",palette = "Set2")+
                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))

@@

The rise in imported items is clearly reflected in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (see Figure \@ref(fig:imports)). Yet beyond conforming to our prior inferences about the LPG-EG periods, the deposition of these imported wares may also cast light on the original pattern drawn out in the present analysis: the late surge in burial activity among some select 9th and 10th century tombs.

@@

In Figure \@ref(fig:importstombcenturies), we see that the tombs already built by the end of the PGB period, particularly those built in the 9th and, to a lesser extent, 10th centuries, were major beneficiaries of the influx of imported items during this time, but that, in the following centuries, in line with the rest of the cemetery, the number of imported items deposited within them declined. Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (such as precious vessels, weapons, jewellery etc.) which, owing to their less developed chronologies, must unfortunately be left out of such diachronic analyses for now. But the beginnings of my final argument are seen here too, namely, that the late surge in burial activity seen across both cemeteries in the late 8th and early 7th centuries would appear to be of a different kind to that of a century earlier.

@@

Figure ... is a boxplot for the aoristic sums of each decade across both cemeteries, colour-coded by ceramic phase. Note that, for clarity, tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively. The important observation is that the PGB and EG exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR). The Orientalising period, on the other hand, exhibit much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs. In other words, while both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was shared somewhat more equally among the total burying population. One speaks of increasing plurality, the other of divergence.

@@

This possibility is furthered somewhat by considering the gini coefficient of the distribution of pottery between the tombs. Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unevenly distributed that material was. And what we see is that, though values were always relatively high (i.e. unequal) within the cemeteries, the major spike in the gini coefficient seen in the Orientalising period contrasts markedly with the modest rise and plateau of the PGB-EG. The gini coefficient is most commonly used in the assessment of income inequality, but I want to stress here that it is not "wealth" in any meaningful sense that we are talking about here. Indeed, if anything, the late flourit of burial activity here discussed may be a case of quantity, and not quality.

@@

Figure ... shows the composition of the total pottery assemblage from each cemetery, and from the seven tombs drawn out earlier, through the centuries. In all cases, there appears to be the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel, a jug, and a drinking vessel. The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters, lekanai and more - declines to become only a minor part of the standardised repertoire of the 7th century. The seven tombs that underpinned the dramatic increase in burial activity in the later 8th century did not break with this mould, and their assemblages seem as conventional as any other during the last century of use in the cemeteries.

It has no doubt been apparent that the main focus of this article has been on pottery and not other artefacts of metal, stone, glass, and organic materials. This is largely because the chronologies for these artefacts are insufficiently developed to allow for the kind of diachronic analysis made thus far. But I do want to briefly comment on one aspect of the distribution of non-ceramic artefacts that may make more sense in the light of the foregoing discussion. There is, perhaps unsurprisingly, a relatively strong (Spearman's rank correlation of `r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most unique and precious objects were recovered from among the larger tombs. But, the monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.

@@

Figure ... thus supports the view which has emerged from the preceeding analysis, that the major increase in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial. To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around the same time. We have seen how this surge in acivity occurred against the backdrop of a declining number of tombs dating back to the 10th and 9th centuries, and I have suggested that the rates of change are too rapid, and unevenly distributed, to be explicable through population growth alone. We have seen how little about the physical dimensions of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasts in the comparative diversity of its associated assemblages, and their distribution across the cemetries. The later burials were marked by great uniformity in the associated grave goods, and this apparently limited and consistent repertoire may explain why many of the most used tombs prove exceptional only in the absolute quantity of artefacts, and not the accoutrements of each individual burial.

@@

@@

>>>>

Looking at Figure \@ref(fig:dimensionsboxplots), though, we can see that, through the centuries, the dimensions of the tombs varied, and sometimes consistently, as with the trend toward narrower *dromoi*. The 9th-century tombs, including all but one of those seven noted above, do not particularly stand out. It does seem that chambers were, on average, larger in this period as well as varying more in size than in any other period. Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on the construction of the larger chambers would have been limited. The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this. It appears at any rate that they were gradually declining in size, albeit with some notable outliers.

@@

imitations<-subset(pots,pots$imitation!="")
imitation_aorist<-colSums(imitations[27:76])
imitation_aorist<-data.frame(imitation_aorist)
colnames(imitation_aorist)<-c("aoristic_sum")
row.names(imitation_aorist)<-dates
setDT(imitation_aorist, keep.rownames = "decade")
imitation_aorist$decade<-as.numeric(imitation_aorist$decade)

ggplot()+geom_col(data=import_aorist,aes(decade,aoristic_sum),alpha=0.2)+
          geom_col(data=allimport_aorist,aes(decade,aoristic_sum, fill=origin),position="dodge")+
            geom_col(data=imitation_aorist,aes(decade,aoristic_sum))+
            theme_bw()+theme(text=element_text(family="Garamond", size=12))+
                labs(x="Date (Years BC)", y="Aoristic Sum")+
                  scale_fill_brewer(name="Origin",palette = "Set2")+
                   scale_y_continuous(expand=c(0,0),limits=c(0,20))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))

@@

The rise in imported items is clearly seen in the tomb assemblages, with a succession of increased Greek and Near Eastern imports respectively in the late 9th and early 8th centuries (see Figure \@ref(fig:imports)). Yet beyond conforming to our prior inferences about the LPG-EG periods, the deposition of these imported wares may also cast light on the original pattern drawn out in the present analysis: the late surge in burial activity among some select 9th and 10th century tombs.

@@

In Figure \@ref(fig:importstombcenturies), we see that the tombs already built by the end of the PGB period, particularly those built in the 9th and, to a lesser extent, 10th centuries, were major beneficiaries of the influx of imported items during this time, but that, in the following centuries, in line with the rest of the cemetery, the number of imported items deposited within them declined. Imported ceramics are used here as something of a proxy for items of status value, a shorthand that would ideally be bolstered by other artefacts (such as precious vessels, weapons, jewellery etc.) which, owing to their less developed chronologies, must unfortunately be left out of such diachronic analyses for now. But, nonetheless, this pattern hints towards a possibility that shall now be explored, that the surge in burial activity seen across both cemeteries in the late 8th and early 7th centuries would appear to be of a different kind to that of a century earlier.

@@

Figure \@ref(fig:bigbox) is a boxplot for the aoristic sums of each decade across both cemeteries, colour-coded by ceramic phase. Note that, for visual clarity, the outliers tomb G in the EG period, and tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively. The important observation is that the PGB and EG exhibit the largest interquartile ranges (IQRs) of any period, and each possess only a single outlier (beyond 1.5 times the IQR). The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers far and above the range of the other tombs. In other words, while both the PGB-EG and EO-LO periods witnessed an expanding volume and range of burial activity, in the earlier period this was distributed somewhat more evenly across the total burying population. One speaks of increasing plurality, the other of divergence.

@@

This possibility is supported further by considering the gini coefficient of the distribution of pottery between the tombs. Using the aoristic sums as a probablistically-weighted estimate of the volume of material deposited in each tomb in each decade, use of the gini coefficient can give an (albeit rough) estimate of how unequally distributed that material was. And what we see is that, though values were always relatively high (i.e. unequal) within the cemeteries, the major spike in the gini coefficient seen in the Orientalising period contrasts markedly with the modest rise and plateau of the PGB-EG. The gini coefficient is most commonly used in the assessment of income inequality, but I want to stress here that it is not "wealth" in any meaningful sense that we are talking about here. Indeed, if anything, the late flourit of burial activity here discussed may be a case of quantity, and not quality.

@@

Figure \@ref(fig:comps) shows the composition of the total pottery assemblage from each cemetery, and from the seven tombs drawn out earlier, through the centuries. In all cases, there appears to be the steady crystallisation of a basic 'burial kit', comprising a pithos and lid, an oil vessel, a jug, and a drinking vessel. The proportion of grave goods comprising 'other' vessels - pyxides, askoi, trays, stands, dinoi, kernoi, kraters, lekanai and more - declines to become only a minor part of the standardised repertoire of the 7th century. The seven tombs that underpinned the dramatic increase in burial activity in the later 8th century did not break with this mould, and their assemblages seem as conventional as any other during the last century of use in the cemeteries.

It has no doubt been apparent that the main focus of this article has been on pottery and not other artefacts of metal, stone, glass, and organic materials. This is largely because the chronologies for these artefacts are insufficiently developed to allow for the kind of diachronic analysis made thus far. But I do want to briefly comment on one aspect of the distribution of non-ceramic artefacts that may make more sense in the light of the foregoing discussion. There is, perhaps unsurprisingly, a relatively strong (Spearman's ρ=`r spearmanobjectspots`) correlation between the number of ceramic and non-ceramic finds across the tombs, and, certainly, many of the most unique and precious objects were recovered from among the larger tombs. But, the monopoly those tombs would appear to have over such objects, perhaps tempting one initially to speak of 'elite' or 'high-status' tombs, evaporates when the number of those objects is divided by the number of burials represented.

@@

Figure \@ref(fig:hists) thus supports the view which has emerged from the foregoing analysis, that the major increase in burial activity at both Fortetsa and the KNC was not a phenomenon linked in any straightforward way to simple principles of 'elite' or 'wealthy' burial. To summarise, towards the end of 8th century BC, a select group of tombs, all of which had been constructed back in the 9th or 10th, became the foci of a major increase in burials, something which did not occur among the tombs built around that time. We have seen how this surge in acivity occurred against the backdrop of a declining number of older tombs in use, and I have suggested that the rates of change are too rapid, and unevenly distributed, to be explicable through population growth alone. We have seen how little about the physical dimensions or early use of these tombs presaged their later popularity, and that, while a similar, if smaller, increase in burial activity occurred around a century earlier, in the PGB-EG periods, this contrasted in the comparative diversity of its associated assemblages, and their distribution across the tombs. The later burials were marked by great uniformity in the associated grave goods, and this apparently limited and consistent repertoire may explain why many of the most used tombs prove exceptional only in the absolute quantity of artefacts, and not the accoutrements of each individual burial.

@@

The preceding analysis has revealed a number of potentially significant patterns in the use of the Fortetsa and Knossos North Cemeteries, in particular the exceptional trajectories of a select group of 10th and 9th century tombs. It only remains to consider how we whould characterise the burying groups responsible for these observed trends. The traditional view is that such groups were immediate families (Brock 1957, 41; Boardman 1967, 63; Coldstream & Huxley 1999; 291). The relatively short lifespan of many tombs (often only two or three generations) has been taken as evidence for direct, lineal inheritance, with the tomb going out of use as families moved, individuals founded new tombs, or the line came to an end. The idea of lineal groups is supported by the, admittedly circumstantial, evidence of men, women and children being found within single tombs, and one (contentious) instance of an apparently hereditary mandibular deformity in three non-contemporary skulls in the same tomb (Musgrave 1996).

The model of lineal inheritance has been recently challenged by Kotsonas (2011, 131), however, who considers it ‘probably the greatest factoid in the archaeology of Crete of the given period’. He argues that the rise in interments during the 8th century, noted also in the present study, is too great to be explained by population growth, as would be necessary for the possibility of direct lineal inheritance to be preserved. Certainly, it seems illogical that population increases would lead to the consolidation of burials within only extant tombs, rather than increasing tomb construction. The trend instead must reflect choices made by the burying groups themselves, which Kotsonas feels may have comprised various elements, including perhaps age- and sex-grades. But it seems probable that such corporate groups, if they existed intersected with, and existed alongside, real or fictive kinship networks.

Yet it is hard to be conclusive, as our current understanding of the social organisation of the early Cretan *poleis* is patchy at best. Perlman and Gagarin's [-@Gagarin2016] recent study of the Cretan law codes reminds us, importantly, if lamentably, that there is an often underappreciated degree of diversity apparent from the Archaic inscriptions in the offices, institutions, and social groups of the early city-states. In the traditional view, political authority was concentrated in the hands of the *kosmoi*, aristocratic magistrates selected from the leading, hereditary *startoi*, or ‘clans’, networks of extended lineages controlling the ancestral estates on which worked the serfs and/or slaves who made up the lower rungs of the social hierarchy (Willetts 1982, 240). But despite elements of this account probably being accurate, others remain equivocal, such as the nature of the *pyla*, or 'tribe', and its relation to the *startos* (which may itself be a military grouping), the distinction between an *astos* and a *poliatas*, both clearly some kind of formulation of the rights-bearing civic actor, and the ancestry and spread of the *agelai*, apparently an adolescent age-grade so far only attested at Eltynia [@Gagarin2016, 36; 78-80]. Indeed, many terms which might illuminate social divisions and groupings within the Archaic Cretan populace are only attested at one or two sites, and even then sometimes appear to differ in their form or function. Frustrating though this can be, it has also seen the image of Cretan austerity and uniformity begin to be challenged, or at least nuanced, by an appreciation of the regional variety and idiosyncracy apparent from the laws.

It wouldn't be helpful at present, then, to draw any simple parallel between the Knossian burial groups and, say, powerful *startoi*, a term which is not even attested outside of Gortyn in this period (ibid., 1-2). But archaeological and textual evidence suggests that, in general terms, such supra-familial kinship groups were undergoing a transition during this period, in ways reflected at Knossos, and thus supporting the attribution of the chamber tombs to such groups. Elsewhere on Crete, changes around this time include a shift to unmarked, individual burials, and more austere grave goods, as well as an apparent reorientation of sanctuary deposits away from extra-mural to suburban spaces, and a rise in utilitarian, mass-produced votives (Perlman & Gagarin 2016, 34-6). At Azoria, major urban remodelling in the 7th century saw the earlier, agglutinative house forms of neighbouring Kavousi Vronda and Kastro give way to larger, more formally planned dwellings with little evidence of renovation or remodelling, while larger, apprently civic, buildings were constructed to accommodate major public gatherings. At Knossos, a small number of pithos burials in pits do appear in the KNC and Fortetsa volumes (Brock 1957, 98; Coldstream & Catling 1996, 162), and recent re-excavation of the Teke plot identified numerous small pit burials around three chamber tombs (Evely 2012), raising the possibility that many more such burials filled the interstices between larger chamber tombs elsewhere, but have gone unrecognised. The insecure dates of these pits, however, makes it unclear whether they represent a genuine Archaic trend.

These varied lines of evidence may all suggest that with the emergence of the archaic *poleis* on Crete, there was a concomitant restructuring of wider social groups, institutions, and behaviours. Perlman and Gagarin (Gagarin & Perlman 2016, 32-7) argue that, at least by the Classical period, supra-household kinship units such as the *startos* or *pyla*, whatever their exact formulations, had become mostly ceremonial, political entities, and had little influence over the social organisation of the Cretan city states. Was this transformation linked to an emerging emphasis on the citizen, the state, and ideologies of relative egality, within which communal gatherings, intra-mural ritual, and individual burial replaced ostentatious, kin-centred acts of gathering, dedication, and burial? It is plausible, though as Perlman and Gagarin (32-37) concede, the decline of ostentatious burial could just as well represent a voluntary abandonment by groups assured of their economic and political influence, finding the symbolic expression of kin-relations in death no longer a necessary expense.

These issues, naturally, recall the conclusions of Morris (1987, 171-211), who argued 30 years ago that the emergence of the *polis* across the Aegean led to a tension between elite preferences for conspicuous display, and a downplaying of social difference in death, as promoted by the ‘middling’ ideology that the *polis* engendered and promulgated. In many areas, this resulted in increasing restriction of elaborate tombs, despite a ‘rise in aristocratic family consciousness’ that saw certain wealthy tombs buck the trend, until ‘such overt groupings were subsumed within the larger “citizen cemeteries”’ (ibid., 185). In his well-known model of the *agathoi* and *kakoi*, Morris argued that, at various times, the relevant spheres of kinship demonstrated in burial could widen and constrict, with more and less distantly related lineages drawn into association through changing social strategies, until the pre-eminence of kinship relations in the structuring of political power declined.

Returning to Knossos, while aspects of the noted expansion in the burying population would seem readily explicable within a model like Morris', certain observations sit more uneasily. Firstly, the burials at issue here are in many ways *not* ostentatious in comparison to the norm. We have seen that the 'richest' tombs in terms of finds are not so 'rich' after all, when one takes account of the number of burials. They are not phyiscally larger, and do not exhibit monopolies over any of a range of potentially 'valuable' offerings. Likewise, Morris claims that periods of expansion in the rite of burial will correspond to greater variety in the associated assemblages, yet, at Knossos, we have seen that the range of grave goods in fact diminished and crystallised during the late 8th and early 7th centuries. Perhaps the expansion was driven by a widening of access to the burial rite by the so-called *agathoi*, with the standardised burial assemblages coming to reflect their shared affinal relations. Or else, the widening of access to these tombs actually heralds the breakdown in traditional kin-based stewardship thereof, perhaps with new, non-lineal corporate groups becoming  the custodians of the tombs, with the limited funereal repertoire thus reflecting the levelling ideology of the emergent state. The exclusive focus of the expansion in a very limited number of older tombs, though, would seem to argue in favour of the former. What seems certain is that this was a period of transition wherein, whatever their identity, certain actors invested heavily in a final flurry of chamber tomb burials, with an emphasis on tradition, an established repertoire of items, rites, and venues for one's passage to the afterlife.

And this image ties into the broader picture developing of the island at this time. It is telling that the peak for imported items in the tombs came in the 9th century; the 'Orientalising' as a time period feels increasingly misapplied to Crete, when, at Knossos at least, it witnessed far lower rates of oriental imports, and only limited adoption of the figural motifs which characterise its appearance elsewhere. The observed rise in imitations of Near Eastern vessels in the late 8th and early 7th century further accords with our growing awareness of the selective nature of archaic and classical Crete's external relations. Erickson has, contrary to many, argued that *symposia* were practised on Crete, but that their practitioners simply favoured simpler vessel forms. The local, selective adoption of, and responses to, wider social and material trends across the contemporary Aegean is perhaps becoming the hallmark of archaic Crete, and finds further justification in the trends identified in the EIA cemeteries at Knossos.

@@

It is important that the present study was a comparative investigation of both the Fortetsa and Knossos North Cemetery. It is notable how similar the observed trends are across both these cemeteries, adding significant credence to the inferences drawn. That both underwent similar developments during the EIA suggests not only that they catered to similar sub-sections of the Knossian population, but that some of the socio-economic and political changes inferred and discussed were real, and community-wide. We should be cautious in speculating as to whether, for instance, particular families, lineages, groups, or classes made persistent use of these different necropoleis, but such possiblities are not inconceivable. Further work should seek to extend these comparisons, both into trends unexplored in the present study (such as, for instance, the curious reuse of a number of 11th century tombs in the late 8th and early 7th centuries), and out to the wider funerary landscape of Knossos. I hope to make just such a study but, for the time being, the datasets I am making available alongside this article include the other published EIA tombs in the area, so that others may also explore their possible significance.

There were undeniable limitations to the present study, also. For one, non-ceramic artefacts were underrepresented, mostly because they lack accurate dating in the manner of pottery and future work could focus on dating these artefacts more closely, permitting their future inclusion into temporal analyses. Issues of gender were sadly also little considered, though this was largely due to limitations in the data themselves. There have also been a number of tombs discovered in the last 20 years that fill in the spaces between the Teke plot and the Medical Faculty Site, and ideally these unpublished tombs should be included in further analyses as they, along with the unconsidered Teke tholos and associated tombs, are almost surely part of a single contiguous cemetery (Blackman 1999, 113; Catling 1983, 51; Evely 2012; French 1991, 68–9). Finally, a preliminary study was recently taken by the author of select boxes marked as 'unpublished' from the KNC deposits kept at the British School at Athens at Knossos. Despite going unmentioned in the publication, this material was, at least in the boxes seen, not inconsiderable and so, regretably, future study will be necessary to determine whether it threatens the integrity of quantitative analyses such as this. Suffice to say that, at present, it is another pinch of salt with which to take the findings.

These limitations notwithstanding, the foregoing analysis demonstrates the value of quantitative approaches to the archaeological record of the period. Such methods allow us not only to identify types of variability in the archaeological material, but also assess the scale of that variability. With large datasets, such as those used in the present study, we are able to locate and visualise covariance and significant correlations between different variables, plot changes in the data through time and, ultimately, draw on our findings to inform discussions of social processes. Though the construction of databases can be time consuming, once made they offer the opportunity to subdivide, compare, and plot the data in numerous ways quickly and easily, facilitating investigation of the material from multiple angles. Such methods are also more independently verifiable than conclusions drawn from more qualitative analyses, for two main reasons. Firstly, though explanations for any patterning in the data still require interpretation, the existence of such patterns may be more easily and reliably assessed via the use of quantitative and statistical tests of correlation and covariance. And secondly, by making datasets available for use by others, as with the data and code for the present study, one’s conclusions can be independently tested, and alternative interpretations can be forwarded, by other scholars. It is hoped that the present study demonstrates the wealth of insights yet to be uncovered by the application of more systematic, quantitative methods. The databases produced for the above analysis contain a large of amount of unutilised data that could facilitate a range of future investigations, and all are welcome to it. With more such analyses, we could begin to a construct a fuller account of the regional diversity in mortuary, and wider social, practices that characterised the communities of EIA and archaic Crete.
